Evolution of pre-copulatory and post-copulatory inbreeding behaviour: a computational

### 2016-09-19 ms_pre_post.Rmd
Update Results and Discussion for new results that very long term evolution of strategies goes to random mating for the relatively more costly strategy.

Figure 1 shows changes in mean allele values of tendency for polyandry, inbreeding mating strategy, and inbreeding fertilisation strategy across all eight possible combinations of costs. In Figure 1A, when no phenotypes are costly, inbreeding mating strategy and inbreeding fertilisation strategy evolve to inbreeding avoidance ($M_{a}<0$ and $F_{a}<0$), and females evolve to be polyandrous ($P_{a}>0$). Costs of phenotypes have a direct effect on evolution, causing inbreeding mating strategy and inbreeding fertilisation strategy phenotypes to evolve to values closer to zero (because realisation of inbreeding strategy costs increases with phenotype absolute values) and tendency for polyandry phenotypes to evolve to values that are more negative (because realisation of polyandry costs increases with positive phenotype values).

We can use Figure 1 to test the hypothesis that inbreeding mating strategy will evolve to inbreeding avoidance ($M_{a}<0$), but inbreeding fertilisation strategy will be negligible ($F_{a} \approx 0$) when mating costs are low ($c_{M}=0$) relative to the costs of tendency for polyandry and inbreeding fertilisation strategy ($c_{P}=0.02$ or $c_{F}=0.02$) by comparing different cost combinations to the null model in which all costs equal zero (Figure 1A). Figures 1A,C,D,G show that inbreeding avoidance evolves at a comparable rate whenever $c_{M}=0$. However, inbreeding fertilisation strategy is never negligible; $F_{a}$ values are still negative over generations even when costs of inbreeding mating strategy is weak and costs of tendency for polyandry and inbreeding fertilisation strategy are strong.

We can also use Figure 1 to test the hypothesis that inbreeding fertilisation strategy will evolve to inbreeding avoidance ($F_{a}<0$), but inbreeding mating strategy will be negligible ($M_{a} \approx 0$) when tendency for polyandry and inbreeding fertilisation strategy costs are low ($c_{P}=0$ and $c_{F}=0$) relative to the cost of inbreeding mating strategy ($c_{M}=0.02$). Figure 1B shows that inbreeding avoidance through fertilisation bias evolves ($F_{a}<0$) under these conditions, but evolution of inbreeding mating strategy is not negligible; $M_{a}$ values are consistently negative across all combinations of costs.

@@

Figure 3 shows evolution of tendency for polyandry and inbreeding fertilisation strategy when inbreeding mating strategy is fixed so that females must mate randomly (blue line at $M_{a}=0$). For all cost combinations of tendency for polyandry and inbreeding fertilisation strategy (Figure 3 panels A-D), we observed selection for positive values of tendency for polyandry alleles and negative values of inbreeding fertilisation strategy alleles, which we interpret as strong evidence for evolution of polyandry and associated inbreeding avoidance through fertiisation bias. This pattern of evolution contrasts with that observed when inbreeding mating strategy was not fixed, indicating that constraints on evolution of inbreeding mating strategy increased the strength of selection for evolution of polyandry and inbreeding avoidance through fertilisation bias. For example, in Figures 1C,D,G,H, tendency for polyandry allele values are typically less than zero, reflecting evolution of monandry, and inbreeding fertilisation strategy allele values were not as negative as they were in simulations with fixed random mating (Figure 3). In general, when females were unable to avoid mating with relatives given severe inbreeding depression ($\beta=3$), selection for evolution of polyandry and post-copulatory inbreeding avoidance was strengthened.

Figure 2, which we previously contrasted with Figure 1 to test whether or not evolution of inbreeding fertilisation strategy facilitates selection for polyandry, shows evolution of tendency for polyandry and inbreeding mating strategy when inbreeding fertilisation strategy is fixed such that the sire of a focal female's offspring is randomly assigned from her mates according to a fair raffle. In all cost combinations of Figure 2 (panels A-D), tendency for polyandry allele values decreased over generations, reflecting evolution of monandry. However, values of inbreeding mating strategy alleles did not greatly differ over generations when inbreeding fertilisation strategy was fixed, meaning that the effect of fixed inbreeding fertilisation strategy on increasing selection for inbreeding avoidance through mate choice was weak or nelgigible. A weak effect is discernable in the contrast between Figure 2B and Figure 1B, in which the grand mean inbreeding mating strategy allele value is $-0.627$ when inbreeding fertilisation strategy is not fixed (Figure 1B) versus $-1.036$ when it is fixed (Figure 2B) in generation 10000. Because females have 20 alleles with additive effects on inbreeding mating strategy, an average female is therefore expected to have mean inbreeding mating strategy phenotype values of $M_{p} = -0.627 \times 20 = -12.54$ and $M_{p} = -1.036 \times 20 = -20.72$ for not fixed versus fixed inbreeding fertilisation strategy, respectively. These values would cause an outbred focal female to be $2.135$ and $4.180$ times more likely to choose a non-relative than a full sibling as a mate given not fixed versus fixed inbreeding fertilisation strategy, respectively. Therefore, when females were unable to bias fertilisation to avoid inbreeding given severe inbreeding depression ($\beta=3$), selection for evolution of pre-copulatory inbreeding avoidance was strengthened.

@@

A low relative cost for one inbreeding strategy (e.g., pre-copulatory inbreeding avoidance) did not preclude a more costly complementary inbreeding strategy from persisting in a focal population (Figure 1), meaning that evolution of both pre- and post-copulatory inbreeding avoidance might be expected in populations despite cost asymmetry. However, existence of costs strongly affected the evolution of the costly phenotype(s), meaning that quantifying costs is necessary for predicting evolution of phenotypes associated with inbreeding strategies. Quantifying direct costs of phenotypes associated with polyandry and mate choice is empirically challenging [@Pomiankowski1987; @Kokko2003; @Reid2015]. @Pomiankowski1987 categorised four types of costs that are relevant to mate choice, which we suggest are also relevant costs for polyandry, and include costs of elevated risks associated with (1) predation or (2) disease transmission, and costs incurred through loss of time (3) or depletion of energy (4). In our model, we interpreted a cost of tendency for polyandy as an elevated risk of predation as might be incurred while searching for or courting mates [e.g., @Rowe1988; @Rowe1994], a cost of inbreeding mating strategy as a lost of time [i.e., risk of not finding a mate in time due to choosiness; e.g., @Kokko2013], and a cost of inbreeding fertilisation strategy as depletion of energy. While these costs are biologically realistic, different costs will be relevant for different populations and be realised in different ways. For example, increased risk of disease transmission has been observed for highly polyandrous females [@Roberts2015a], but such a cost would more realistically be applied to the number of realised mates a female has rather than her tendency for engaging in polyandry. Future models will therefore need to carefully consider the magnitudes and realisation of direct costs to accurately predict evolution of inbreeding strategies in populations. <!--- add Tuni citation for depletion of energy? --->

>>>>

Figure 1 shows changes in mean allele values of tendency for polyandry, inbreeding mating strategy, and inbreeding fertilisation strategy across all eight possible combinations of costs. In Figure 1A, when no phenotypes are costly, inbreeding mating strategy and inbreeding fertilisation strategy evolve to inbreeding avoidance ($M_{a}<0$ and $F_{a}<0$), and females evolve to be polyandrous ($P_{a}>0$). Costs of phenotypes had a direct effect on evolution, causing inbreeding mating strategy and inbreeding fertilisation strategy phenotypes to evolve to values closer to zero (because realisation of inbreeding strategy costs increases with phenotype absolute values) and tendency for polyandry phenotypes to evolve to values that were more negative (because realisation of polyandry costs increases with positive phenotype values).

We can use Figure 1 to test the hypothesis that inbreeding mating strategy will evolve to inbreeding avoidance ($M_{a}<0$), but inbreeding fertilisation strategy will be negligible ($F_{a} \approx 0$) when mating costs are low ($c_{M}=0$) relative to the costs of tendency for polyandry and inbreeding fertilisation strategy ($c_{P}=0.02$ or $c_{F}=0.02$) by comparing different cost combinations to the null model in which all costs equal zero (Figure 1A). Figure 1A,C,D,G show that inbreeding avoidance in mating ($M_{a}<0$) evolved consistently whenever $c_{M}=0$. When inbreeding fertilisation strategy was costly and inbreeding avoidance strategy was not costly (Figure 1D,G), inbreeding fertilisation strategy initially evolved toward inbreeding avoidance ($F_{a} < 0$), but long-term evolution of inbreeding fertilisation strategy stabilised at random mating ($F_{a} \approx 0$). However, when only tendency for polyandry was costly (Figure 1C), both inbreeding mating strategy and inbreeding fertilisation strategy evolved to inbreeding avoidance over generations ($M_{a} < 0$ and $F_{a} < 0$).

We can also use Figure 1 to test the hypothesis that inbreeding fertilisation strategy will evolve to inbreeding avoidance ($F_{a}<0$), but inbreeding mating strategy will be negligible ($M_{a} \approx 0$) when tendency for polyandry and inbreeding fertilisation strategy costs are low ($c_{P}=0$ and $c_{F}=0$) relative to the cost of inbreeding mating strategy ($c_{M}=0.02$). Figure 1B shows that inbreeding avoidance through fertilisation bias evolved ($F_{a}<0$) under these conditions. While inbreeding mating strategy initially evolved toward inbreeding avoidance ($M_{a} < 0$), long-term evolution of inbreeding mating strategy stabilised at random mating ($M_{a} \approx 0$).

@@

Figure 3 shows evolution of tendency for polyandry and inbreeding fertilisation strategy when inbreeding mating strategy is fixed so that females must mate randomly (blue line at $M_{a}=0$). For all cost combinations of tendency for polyandry and inbreeding fertilisation strategy (Figure 3 panels A-D), we observed selection for positive values of tendency for polyandry alleles and negative values of inbreeding fertilisation strategy alleles, which we interpret as strong evidence for evolution of polyandry and associated inbreeding avoidance through fertiisation bias. This pattern of evolution contrasts with that observed when inbreeding mating strategy was not fixed, indicating that constraints on evolution of inbreeding mating strategy increased the strength of selection for evolution of polyandry and inbreeding avoidance through fertilisation bias. For example, in Figures 1C,D,G,H, tendency for polyandry allele values were typically less than zero, reflecting evolution of monandry, and inbreeding fertilisation strategy allele values were not as negative as they were in simulations with fixed random mating (Figure 3). In general, when females were unable to avoid mating with relatives given severe inbreeding depression ($\beta=3$), selection for evolution of polyandry and post-copulatory inbreeding avoidance was strengthened.

Figure 2, which we previously contrasted with Figure 1 to test whether or not evolution of inbreeding fertilisation strategy facilitates selection for polyandry, shows evolution of tendency for polyandry and inbreeding mating strategy when inbreeding fertilisation strategy is fixed such that the sire of a focal female's offspring is randomly assigned from her mates according to a fair raffle. In all cost combinations of Figure 2 (panels A-D), tendency for polyandry allele values decreased over generations, reflecting evolution of monandry. However, values of inbreeding mating strategy alleles did not greatly differ over generations when inbreeding fertilisation strategy was fixed, meaning that the effect of fixed inbreeding fertilisation strategy on increasing selection for inbreeding avoidance through mate choice was weak or nelgigible. A weak effect is discernable in the contrast between Figure 2B and Figure 1B, in which the grand mean inbreeding mating strategy allele value is $-0.627$ when inbreeding fertilisation strategy is not fixed (Figure 1B) versus $-1.036$ when it is fixed (Figure 2B) in generation 10000. Because females have 20 alleles with additive effects on inbreeding mating strategy, an average female was therefore expected to have mean inbreeding mating strategy phenotype values of $M_{p} = -0.627 \times 20 = -12.54$ and $M_{p} = -1.036 \times 20 = -20.72$ for not fixed versus fixed inbreeding fertilisation strategy, respectively. These values would cause an outbred focal female to be $2.135$ and $4.180$ times more likely to choose a non-relative than a full sibling as a mate given not fixed versus fixed inbreeding fertilisation strategy, respectively. Therefore, when females were unable to bias fertilisation to avoid inbreeding given severe inbreeding depression ($\beta=3$), selection for evolution of pre-copulatory inbreeding avoidance was strengthened.

@@

A low relative cost for one inbreeding strategy (e.g., pre-copulatory inbreeding avoidance) precluded a more costly complementary inbreeding strategy from persisting in a focal population (Figure 1), meaning that long-term persistence of both pre- and post-copulatory inbreeding avoidance might not be expected in populations given cost asymmetry. Direct costs therefore strongly affected the evolution of the costly phenotype(s), meaning that quantifying costs is necessary for predicting evolution of phenotypes associated with inbreeding strategies. Quantifying direct costs of phenotypes associated with polyandry and mate choice is empirically challenging [@Pomiankowski1987; @Kokko2003; @Reid2015]. @Pomiankowski1987 categorised four types of costs that are relevant to mate choice, which we suggest are also relevant costs for polyandry, and include costs of elevated risks associated with (1) predation or (2) disease transmission, and costs incurred through loss of time (3) or depletion of energy (4). In our model, we interpreted a cost of tendency for polyandy as an elevated risk of predation as might be incurred while searching for or courting mates [e.g., @Rowe1988; @Rowe1994], a cost of inbreeding mating strategy as a lost of time [i.e., risk of not finding a mate in time due to choosiness; e.g., @Kokko2013], and a cost of inbreeding fertilisation strategy as depletion of energy. While these costs are biologically realistic, different costs will be relevant for different populations and be realised in different ways. For example, increased risk of disease transmission has been observed for highly polyandrous females [@Roberts2015a], but such a cost would more realistically be applied to the number of realised mates a female has rather than her tendency for engaging in polyandry. Future models will therefore need to carefully consider the magnitudes and realisation of direct costs to accurately predict evolution of inbreeding strategies in populations. <!--- add Tuni citation for depletion of energy? --->
